---
title: "Florida Census Demographics (2000–2020)"
---

```{r}
#| label: setup
#| message: false
#| echo: false

library(tidyverse)
library(sf)
library(leaflet)
library(htmltools)
library(htmlwidgets)

# Load county boundaries and decennial census data
fl_geo  <- read_rds("data/fl_geo.rds")
fl_race <- read_rds("data/fl_race_2000_2020.rds")

# Reshape data to wide format to compute percentage point changes (2000 to 2020)
race_wide <- fl_race |>
  mutate(pct = round(100 * (value / summary_value), 1)) |>
  select(GEOID, variable, year, pct) |>
  pivot_wider(names_from = c(variable, year), values_from = pct) |>
  mutate(
    diff_hisp  = round(hispanic_2020 - hispanic_2000, 1),
    diff_black = round(black_2020 - black_2000, 1),
    diff_white = round(white_2020 - white_2000, 1),
    diff_asian = round(asian_2020 - asian_2000, 1)
  )

# Combine with county geometries and reproject to WGS84
fl_map <- left_join(fl_geo, race_wide, by = "GEOID") |>
  st_transform(4326)

# Create race-specific hover labels for each county
labels_hisp <- sprintf(
  "<strong>%s</strong><br/>Hispanic: %s%% &rarr; %s%% (%+0.1f pts)",
  fl_map$NAME, fl_map$hispanic_2000, fl_map$hispanic_2020, fl_map$diff_hisp
) |> lapply(HTML)

labels_white <- sprintf(
  "<strong>%s</strong><br/>White: %s%% &rarr; %s%% (%+0.1f pts)",
  fl_map$NAME, fl_map$white_2000, fl_map$white_2020, fl_map$diff_white
) |> lapply(HTML)

labels_black <- sprintf(
  "<strong>%s</strong><br/>Black: %s%% &rarr; %s%% (%+0.1f pts)",
  fl_map$NAME, fl_map$black_2000, fl_map$black_2020, fl_map$diff_black
) |> lapply(HTML)

labels_asian <- sprintf(
  "<strong>%s</strong><br/>Asian: %s%% &rarr; %s%% (%+0.1f pts)",
  fl_map$NAME, fl_map$asian_2000, fl_map$asian_2020, fl_map$diff_asian
) |> lapply(HTML)

# Define color scales for each demographic group
pal_hisp  <- colorNumeric("YlOrRd", domain = fl_map$diff_hisp)
pal_white <- colorNumeric("Blues", domain = fl_map$diff_white)
pal_black <- colorNumeric("Purples", domain = fl_map$diff_black)
pal_asian <- colorNumeric("Greens", domain = fl_map$diff_asian)
```

## Two Decades of Demographic Change Across Florida

Between 2000 and 2020, Florida saw significant demographic shifts across all major racial and ethnic groups:

- **Hispanic**: Substantial increases statewide, especially in central and southwest Florida.
- **White (non-Hispanic)**: Decreased as a overall percentage of population in almost all 67 counties.
- **Black**: Steady population shares in northern counties, with growth in select central and southern metros.
- **Asian**: Steady percentage growth across major suburban and university counties.

Use the radio buttons in the upper right to toggle which demographic shift colors the map. Hover over any county to view the 20-year shift for the selected group.

```{r}
#| label: interactive-map
#| message: false
#| echo: false

leaflet(fl_map, width = "100%", height = "540px") |>
  addTiles() |>
  addPolygons(
    group = "Hispanic Change",
    fillColor = ~pal_hisp(diff_hisp),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels_hisp
  ) |>
  addPolygons(
    group = "White Change",
    fillColor = ~pal_white(diff_white),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels_white
  ) |>
  addPolygons(
    group = "Black Change",
    fillColor = ~pal_black(diff_black),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels_black
  ) |>
  addPolygons(
    group = "Asian Change",
    fillColor = ~pal_asian(diff_asian),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels_asian
  ) |>
  addLayersControl(
    baseGroups = c("Hispanic Change", "White Change", "Black Change", "Asian Change"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(
    pal = pal_hisp, values = ~diff_hisp,
    title = "Hispanic (% pts)",
    className = "info legend legend-hisp",
    position = "bottomright"
  ) |>
  addLegend(
    pal = pal_white, values = ~diff_white,
    title = "White (% pts)",
    className = "info legend legend-white",
    position = "bottomright"
  ) |>
  addLegend(
    pal = pal_black, values = ~diff_black,
    title = "Black (% pts)",
    className = "info legend legend-black",
    position = "bottomright"
  ) |>
  addLegend(
    pal = pal_asian, values = ~diff_asian,
    title = "Asian (% pts)",
    className = "info legend legend-asian",
    position = "bottomright"
  ) |>
  onRender("
    function(el, x) {
      var map = this;
      function updateSelection(activeLayer) {
        // Hide all legends and plots
        document.querySelectorAll('.legend-hisp, .legend-white, .legend-black, .legend-asian').forEach(function(l) {
          l.style.display = 'none';
        });
        document.querySelectorAll('.race-plot').forEach(function(p) {
          p.style.display = 'none';
        });

        // Show matching legend and plot
        if (activeLayer === 'Hispanic Change') {
          document.querySelectorAll('.legend-hisp').forEach(function(l) { l.style.display = 'block'; });
          var p = document.getElementById('plot-hisp');
          if (p) p.style.display = 'block';
        } else if (activeLayer === 'White Change') {
          document.querySelectorAll('.legend-white').forEach(function(l) { l.style.display = 'block'; });
          var p = document.getElementById('plot-white');
          if (p) p.style.display = 'block';
        } else if (activeLayer === 'Black Change') {
          document.querySelectorAll('.legend-black').forEach(function(l) { l.style.display = 'block'; });
          var p = document.getElementById('plot-black');
          if (p) p.style.display = 'block';
        } else if (activeLayer === 'Asian Change') {
          document.querySelectorAll('.legend-asian').forEach(function(l) { l.style.display = 'block'; });
          var p = document.getElementById('plot-asian');
          if (p) p.style.display = 'block';
        }
      }
      updateSelection('Hispanic Change');
      map.on('baselayerchange', function(e) {
        updateSelection(e.name);
      });
    }
  ")
```

## Top Counties by Demographic Shift

Select a demographic group in the map above to view the top 10 counties with the largest shift below.

::: {#plot-hisp .race-plot}
```{r}
#| label: plot-hisp
#| echo: false
#| message: false

fl_map |>
  st_drop_geometry() |>
  slice_max(diff_hisp, n = 10) |>
  mutate(county = fct_reorder(str_remove(NAME, " County, Florida"), diff_hisp)) |>
  ggplot(aes(x = diff_hisp, y = county)) +
  geom_col(fill = "#e6550d") +
  geom_text(aes(label = paste0("+", diff_hisp, " pts")), hjust = -0.15, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Top 10 Florida Counties by Hispanic Share Growth (2000–2020)",
    x = "Percentage Point Increase",
    y = NULL,
    caption = "Source: U.S. Census Bureau (SF1 & PL 94-171)"
  ) +
  theme_minimal()
```
:::

::: {#plot-white .race-plot style="display: none;"}
```{r}
#| label: plot-white
#| echo: false
#| message: false

fl_map |>
  st_drop_geometry() |>
  slice_min(diff_white, n = 10) |>
  mutate(county = fct_reorder(str_remove(NAME, " County, Florida"), desc(diff_white))) |>
  ggplot(aes(x = diff_white, y = county)) +
  geom_col(fill = "#3182bd") +
  geom_text(aes(label = paste0(diff_white, " pts")), hjust = 1.15, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0.18, 0))) +
  labs(
    title = "Top 10 Florida Counties by White Share Decline (2000–2020)",
    x = "Percentage Point Change",
    y = NULL,
    caption = "Source: U.S. Census Bureau (SF1 & PL 94-171)"
  ) +
  theme_minimal()
```
:::

::: {#plot-black .race-plot style="display: none;"}
```{r}
#| label: plot-black
#| echo: false
#| message: false

fl_map |>
  st_drop_geometry() |>
  slice_max(diff_black, n = 10) |>
  mutate(county = fct_reorder(str_remove(NAME, " County, Florida"), diff_black)) |>
  ggplot(aes(x = diff_black, y = county)) +
  geom_col(fill = "#756bb1") +
  geom_text(aes(label = paste0("+", diff_black, " pts")), hjust = -0.15, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Top 10 Florida Counties by Black Share Growth (2000–2020)",
    x = "Percentage Point Increase",
    y = NULL,
    caption = "Source: U.S. Census Bureau (SF1 & PL 94-171)"
  ) +
  theme_minimal()
```
:::

::: {#plot-asian .race-plot style="display: none;"}
```{r}
#| label: plot-asian
#| echo: false
#| message: false

fl_map |>
  st_drop_geometry() |>
  slice_max(diff_asian, n = 10) |>
  mutate(county = fct_reorder(str_remove(NAME, " County, Florida"), diff_asian)) |>
  ggplot(aes(x = diff_asian, y = county)) +
  geom_col(fill = "#31a354") +
  geom_text(aes(label = paste0("+", diff_asian, " pts")), hjust = -0.15, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Top 10 Florida Counties by Asian Share Growth (2000–2020)",
    x = "Percentage Point Increase",
    y = NULL,
    caption = "Source: U.S. Census Bureau (SF1 & PL 94-171)"
  ) +
  theme_minimal()
```
:::

